QTL information page and tutorial

### 2023-12-15 DRAFT_QTL_tutorial.Rmd
progress on QTL page

Initial publication year: 2023 <br>
[How to cite](https://marineomics.github.io/#How_to_Cite)

@@

*I owe most of this code and most of what I know about QTL analysis to Dr. Evangeline Ballerini.

@@

## Tutorial
This tutorial will walk you through the essentials of the particular analysis that I did, but my hope is that it will be useful even for those who take different approaches. It is not written with reproducible data, but I have included all of the necessary code and outputs.

I used whole genome skim sequencing due to the ease of a multiplex library prep, making costs and wet lab time comparable to reduced representation methods.

@@

### We're going to start by defining all of our filepaths, variable names, and tools
```{bash, eval = F}
# make variables for commonly used directories. as always, set your working directory wherever works for you
workingdir="/Users/jason/qtl.tutorial"
mkdir $workingdir

@@

# file with the Aquilegia reference genome, and bams for the f1 and the parent I have sequence for ('Pa')
reference="/Volumes/HD/genomes.reference/Aquilegia/V3/sequences/Aquilegia_coerulea.main_genome.scaffolds.fasta"
f1bam="$bamdir/f2.merged.bam"
pabam="/Volumes/HD/projects/genomes.species/V3/alignments/jonesii.sorted.bam"

@@

### To begin our genotyping, let's determine the useful sites for genotyping the F2s
####  Generate a genotype depth sequence file for the F1
  * Of course, this can be done by sequencing the F1 to enough coverage to confidently genotype it at many SNPs, however there is a much easier and less costly way:
    * In this case, the F2 population only inherited alleles from the F1, so if we merge all of the F2 sequence data, we're left with a genotype level sequence file of the F1!

@@

# make a list file with all of the file paths of our .bams for samtools to merge
ls -d -1 $bamdir/*.* > $bamdir/f2.paths.list

# merge bam files
samtools merge $bamdir/f2.merged.bam -b $bamdir/f2.paths.list

# index bam files
samtools index $bamdir/f2.merged.bam
```

@@

    ```
    * We have our list of high quality, heterozygus sites from the F1, so now we need to determine their genotype in Pa and output that vcf file
      * The below code uses [samtools mpileup](http://www.htslib.org/doc/samtools-mpileup.html) to take particular sites from a .bam file (f1.het.pos) and make a .vcf file from it with [bcftools call](https://samtools.github.io/bcftools/bcftools.html#call). You can look up the individuals commands on the linked manual pages.
        ```{bash, eval = F}
        samtools mpileup -l f1.het.pos -q 30 -u -B -f $reference $pabam | bcftools call -m -f GQ > pa.genotypes.vcf
        ```
    * Now we grab sites homozygous in Pa
      * SNPs only
      * high quality
      * 20x < coverage < 100x (coverage on this parent was ~80x)
      * note that all of our parameters are 'removing'excluding' sites with those specs due to the '-e' in bcftools query (e.g. DP<20 = remove sites with less than 20x coverage). You could do this either way.
        ```{bash, eval = F}
        bcftools filter -g3 pa.genotypes.vcf | awk ' ($8 !~ /INDEL/) {print $0} ' | bcftools query -e 'QUAL<30 || DP<20 || DP>100 || GQ<100 || GT="0/1" || GT="1/2"'      -f'%CHROM\t%POS\n' > snp.list
        ```
    * snp.list is our final list of usable SNPs, of which there were ~1,000,000 in this case
    * Now we'll call genotypes for both the F1 and Pa at these sites using a combination of [samtools mpileup](http://www.htslib.org/doc/samtools-mpileup.html), [bcftools call](https://samtools.github.io/bcftools/bcftools.html#call), and [bcftools query](https://samtools.github.io/bcftools/bcftools.html#query). This is a simple text file with 4 columns: 
      * chromosome, position, reference allele, and the alternate allele:
      ```{bash, eval = F}
      samtools mpileup -l snp.list -q 30 -u -B -f $reference $f1 $pabam | bcftools call -vm -f GQ | awk ' ($8 !~ /INDEL/) {print $0} ' | bcftools query -f        '%CHROM\:%POS\t%REF\t%ALT[\t%GT]\n' > f1.pa.genotypes
      ```
    * Finally, we'll print a file with the sites, the Pa allele, and the Pb allele
    ```{bash, eval = F}
    awk ' { if ( $5 ~ /0\/0/ ) print $1 "\t" $2 "\t" $3; else print $1 "\t" $3 "\t" $2; } ' f1.pa.genotypes > pos.pa.pb.txt    
    ```

### Now we can actually genotype the F2s
 * determine useful window sizes depending on recombination rate
    * we used from 10kb to 1Mb
    * make a .csv or .txt defining these windows
  * calculate frequencies of parental alleles in these windows
  * we used windows of allele frequencies of the parent we had sequence for (Pa) to determine F2 genotypes
    * frequency of Pa allele
      * 0 - 0.15: homozygous Pb
      * \>0.15 - 0.35: unsure
      * \>0.35 - 0.65: heterozygous
      * \>0.65 - 0.85: unsure
      * \>0.85 - 1: homozygous Pa

>>>>

@@

*All of this code and most of what I know about QTL analysis to [Prof. Evangeline Ballerini](https://www.ballerinilab.com/). Add something about Nathan. Any mistakes or lack of clarity are mine.

@@

# Tutorial

This tutorial will walk you through the essentials of the particular analysis that I did, but my hope is that it will be useful even for those who take different approaches. It is not written with reproducible data, but I have included all of the necessary code and outputs.

## Genetic map construction
The first step in QTL mapping is making a genetic map, where we have the determined the genotype of each F2 within recombination windows.

There are many different approaches for doing this, but below I describe our lab's method of using whole genome skim sequencing and custom scripts. I used a multiplexed library prep kit (formerly iGenomx, now Twist 96-plex kit), making costs and wet lab time comparable to reduced representation methods.

@@

### defining all of our filepaths, variable names, and tools
```{bash, eval = F}
# make variables for commonly used directories. as always, set your working directory wherever works for you
workingdir="/Users/jason/qtl.tutorial"
mkdir $workingdir

@@

# file with the Aquilegia reference genome, and bams for the f1 and the parent I have sequence for ('Pa')
reference="/Volumes/HD/genomes.reference/Aquilegia/V3/sequences/Aquilegia_coerulea.main_genome.scaffolds.fasta"
f1bam="$genodir/f2.merged.bam"
pabam="/Volumes/HD/projects/genomes.species/V3/alignments/jonesii.sorted.bam"

@@

### determining the useful sites for genotyping the F2s
####  Generate a genotype depth sequence file for the F1
  * Of course, this can be done by sequencing the F1 to enough coverage to confidently genotype it at many SNPs, however there is a much easier and less costly way:
    * In this case, the F2 population only inherited alleles from the F1, so if we merge all of the F2 sequence data, we're left with a genotype level sequence file of the F1!

@@

# make a list file with all of the file paths of our .bams for samtools to merge
ls -d -1 $bamdir/*.* | grep -v bai > $infodir/f2.paths.list

# merge bam files
samtools merge $genodir/f2.merged.bam -b $genodir/f2.paths.list

# index bam files
samtools index $genodir/f2.merged.bam
```

@@

    ```
    * We have our list of high quality, heterozygus sites from the F1, so now we need to determine their genotype in Pa and output that vcf file
      * The below code uses [samtools mpileup](http://www.htslib.org/doc/samtools-mpileup.html) to take particular sites from a .bam file (f1.het.pos) and make a .vcf file from it with [bcftools call](https://samtools.github.io/bcftools/bcftools.html#call). You can look up the individuals commands on the linked manual pages.
        ```{bash, eval = F}
        samtools mpileup -l f1.het.pos -q 30 -u -B -f $reference $pabam | bcftools call -m -f GQ > pa.genotypes.vcf
        ```
    * Now we grab sites homozygous in Pa
      * SNPs only
      * high quality
      * 20x < coverage < 100x (coverage on this parent was ~80x)
      * note that all of our parameters are 'removing'excluding' sites with those specs due to the '-e' in bcftools query (e.g. DP<20 = remove sites with less than 20x coverage). You could do this either way.
        ```{bash, eval = F}
        bcftools filter -g3 pa.genotypes.vcf | awk ' ($8 !~ /INDEL/) {print $0} ' | bcftools query -e 'QUAL<30 || DP<20 || DP>100 || GQ<100 || GT="0/1" || GT="1/2"'      -f'%CHROM\t%POS\n' > snp.list
        ```
    * snp.list is our final list of usable SNPs, of which there were ~1,000,000 in this case
    * Now we'll call genotypes for Pa at these sites using a combination of [samtools mpileup](http://www.htslib.org/doc/samtools-mpileup.html), [bcftools call](https://samtools.github.io/bcftools/bcftools.html#call), and [bcftools query](https://samtools.github.io/bcftools/bcftools.html#query). This is a simple text file with 4 columns: 
      * chromosome, position, reference allele (which is Pb's allele), the alternate allele (which is Pa's allele), and Pa's genotype, which in this case will always be 1/1 for alternate/alternate:
      ```{bash, eval = F}
      samtools mpileup -l snp.list -q 30 -u -B -f $reference $f1 $pabam | bcftools call -vm -f GQ | awk ' ($8 !~ /INDEL/) {print $0} ' | bcftools query -f        '%CHROM\:%POS\t%REF\t%ALT[\t%GT]\n' > pa.genotypes
      ```
    * Finally, we'll print a file with the site, the Pa allele, and the Pb allele.
    ```{bash, eval = F}
    awk ' { if ( $5 ~ /0\/0/ ) print $1 "\t" $2 "\t" $3; else print $1 "\t" $3 "\t" $2; } ' pa.genotypes > pos.pa.pb.txt    
    ```

### genotyping the F2s
The first step is to go through each F2 individual and count the number of reads of each allele. We'll do this using a combination of samtools mpileup and a custom perl script written by Prof. Anji Ballerini (make sure true). The mpileup command goes through our F2 .bam files and counts alleles for each individual at our snp list. The output has 6 columns:
  * 1: chromosome/scaffold
  * 2: position
  * 3: reference allele
  * 4: read counts at that locus
  * 5: allele
    * . matches reference on forward strand
    * , matches reference on reverse strand
    * uppercase letter is identity of alternate allele read on forward strand and lowercase is the same but on the reverse strand
  * 6: ASCII base quality

I won't pretend to speak perl, but I know this script takes the mpileup as input and outputs a 'counts' file with 5 columns:
  * 1: chromosome/scaffold:position
  * 2: # A reads
  * 3: # T reads
  * 4: # C reads
  * 5: # G reads

Below is the for loop that goes through each F2, calling on mpileup and the perl script to output count files.

```{bash, eval = F}
# make the snp.list a variable that we can easily call
snplist="$idallelesdir/snp.list"

# make the filepath to the perl script a variable
MAYBE LINK TO A NEW HTML WITH THE FILE

# make a list of the F2s from the .bamdir
ls $bamdir | grep -v bai | awk '{ gsub(/.nodup.bam/, "") ; print $0 }' > $infodir/f2.list

#define a list of variables that will be known as "samples" using our f2.list file
samples=($(cat $infodir/f2.list))

# make a for loop that will cycle through the list of F2 samples one at a time and genotype each sample at our SNP.list positions
for (( k = 0 ; k < ${#samples[@]} ; k++ ))
do

samtools mpileup -Q 30 -q 60 -B -f $reference -l $snplist $bamdir/${samples[$k]}.nodup.bam | perl $countscript | awk ' { print $1":"$2 "\t" $3 "\t" $4 "\t" $5 "\t" $6 } ' | sort -k 1,1 > ${samples[$k]}.counts

done
```

Now that we have small and manageable counts files for each F2, we can leave the world of bash and head to R to finish up our genotyping and genetic map construction. Because we're calling genotypes in recombination bins, we need to determine our bin sizes to accurately capture our recombination points. We used anywhere from 0.1 - 1Mb bins, depending on where on the chromosome we were. Generally we used 0.5Mb bins on the ends of chromosome arms and 1Mb bins to account for spatially differing recombination rates. 0.1Mb bins were used in regions that were known to have discordance between the genetic and physical maps based om previous studies in *Aquilegia*.

 * determine useful window sizes depending on recombination rate
    * we used from 10kb to 1Mb
    * make a .csv or .txt defining these windows
  * calculate frequencies of parental alleles in these windows
  * we used windows of allele frequencies of the parent we had sequence for (Pa) to determine F2 genotypes
    * frequency of Pa allele
      * 0 - 0.15: homozygous Pb
      * \>0.15 - 0.35: unsure
      * \>0.35 - 0.65: heterozygous
      * \>0.65 - 0.85: unsure
      * \>0.85 - 1: homozygous Pa
